---
title: "Exam Summary - Group: Data Freak"
author: "Fatemeh, Pratik, Sebastian"
date: "2025-09-26"
output: html_document
---


```{r libraries and data sets, include=FALSE} 
library(tidyverse)
library(here)
library(GGally)
library(ggridges)

data <- read_delim(here("data/raw_data", "exam_data.txt"), delim = "\t")
data_join <- read_delim(here("data/raw_data", "exam_data_join.txt"), delim = "\t")
data_tidy <- read_delim(here("data", "exam_data_tidy_2025-09-08.txt"), delim = " ")
data_tidier <- read_delim(here("data", "exam_data_tidier_2025-09-09.txt"), delim = " ")
```

## Report

This is the exam summary for group `"Data Freak"`, consisting of 3 people. This project is part of the RMED901 course, and the answers were prepared by our group.

We created the GitHub repository `"data_freak_exam_project"` for this exam. This log is made to show that each of us worked on a part of the final report and to be used for the final evaluation.

## Folder structure

The main folder contains a README.md with the complete folder structure and containing files.</br>
The repository contains the following folders:

    scripts:  - one folder for every task 1 - 6, 
              - scripts from each group member for each task, with the corresponding part of the task, 
              - one merged script per task with all important code used for pipelines and/or plots
              - a descriptive .md-file for every task 
              - the exam summary as .Rmd- and .html-file in task 6
              
    data:     - including all 2 used raw data sets, a codebook and 2 created tidy and morphed data sets for use in scripts
    
    results:  - two created tidied and morphed data sets as result for task 2 and 3
              - corresponding codebooks for each data set
              - .html output of the end report for exam
            
    questions:- exam questions


## Raw data

We received a data set split into two parts: `exam_data.txt` and `exam_data_join.txt`, as well as the corresponding codebook (`codebook_exam_data.html`).

`exam_data.txt`:
```{r data, echo=FALSE}
head(data, n = 1L)
```

`exam_data_join.txt`:
```{r data_join, echo=FALSE}
head(data_join, n = 1L)
```

## Tasks

### Task 1 - Create an RStudio project and GitHub repository.

We created an RStudio project and GitHub repository with the name `data_freak_exam_project` and is made in RStudio, with GitHub for collaboration. </br>
URL: `https://github.com/FatemehH-1999/data_freak_exam_project.git`

Folder structure was described above.

### Task 2 - Read and tidy the dataset.

#### Loading libraries, read and glimpse at data set
To read and tidy the given raw data `exam_data.txt` we first loaded the libraries `here` and `tidyverse`. The data set `exam_data.txt` was imported into R using the `read_delim` function with the delimiter `"\t"`.
We glimpsed at the data set, which contains <em>34048</em> observations of <em>15</em> variables. 

#### Checking for duplicates
Next, we checked for duplicates and found <em>12344</em> unique entries. This points to duplicates being present.
    
#### Checking for columns with unfavorable name structure
We checked for difficult for R to interpret column names. We found 5 that need columns to be changed: </br>
`gender-age`, `pan day`, `1_test_id`, `time measurement` and `.value`  -->  </br>
`gender_age`, `pan_day`, `test_id`, `time_measurement` and `value`.

#### Looking for columns to separate
Two columns need to be split up to acquire a tidy data set:</br>
      `subject` -> separated into `ID`, `First_Name`, `Last_Name` and </br>
      `gender_age` -> separated into `gender` and `age`
      
#### Checking variable types
The two columns `ID` and `age` have to be changed from type `character` to type `numeric`.

#### Writing file to save tidied data set
Lastly, we saved the new data set as `exam_data_tidy_2025-09-08.txt` with the function `write_delim` and the delimiter `" "`.
A glimps at the data set, revealed <em>34048</em> observations of now <em>18</em> variables. 

`exam_data_tidy_2025-09-08.txt`:
```{r data_tidy, echo=FALSE}
head(data, n = 1L)
```

### Task 3 - Tidy, adjust, and explore. 

#### Loading libraries, read and glimpse at data set
To read and further tidy our data set `exam_data_tidy_2025-09-08.txt` we first loaded the libraries `here` and `tidyverse`. The data set `exam_data.txt` was imported into R using the `read_delim` function with the delimiter `" "`.

#### Necessary changes in variable types
As described before, two columns needed a change in variable type: `ID` and `age` from `character` to `numeric`, so we converted both variable types.

#### Removing unnecessary columns
We were tasked to remove three columns from our data set: `row`, `test_id`, `demo_group`.

#### Create a set of new columns
Next, we were tasked to create new - or change existing columns:</br>
    - create a column `rec_ver_tat_level` showing, if `rec_ver_tat` is higher than <em>100</em> or not, with values `High/Low`</br>
    - create a numeric column `pan_week`showing `pan_day` in weeks</br>
    - change the column `drive_thru_ind` from `numeric` values <em>1</em> and <em>0</em> to `character` values <em>Yes/No</em></br>
    - create a numeric column `multi_ct_result_orderset` showing multiplication of `ct_result` and `orderset` for each person</br>

#### Change order of columns
We changed the position of columns `ID`, `age`, `gender` to <em>1</em>, <em>2</em> and <em>3</em>, followed by the rest of the colums.

#### Ascending order 
The whole data frame was rearranged by ascending values in column `ID`.

#### Read and join the additional dataset to your main dataset
`exam_data_join.txt` was loaded into R using the `read_delim` function with the delimiter `"\t"`.
A glimpsed at the data set revealed <em>200</em> observations of <em>2</em> variables. 
The two data sets were joined with `left_join` function by column `ID`.

#### Explore and comment on the missing variables
The three columns `payor_group`, `patient_class` and `antibody` contain missing values. 

#### Stratifying of data by each person

##### Sebastian

##### Only for persons tested `pan_day` later than <em>50</em> </br>
Getting every ID/person tested <em>50</em> days after pandemic start, sort after `pan_day`, calculating and outputting min, max, mean and sd of column `age` after grouping by days after pandemic (`pan_day`):

- for each day after start of pandemic (only 4 rows shown)
```{r pan_day_each, echo=FALSE}
age_stratified_after_panday_each <- data_tidier %>% 
  filter(pan_day > 50) %>% 
  group_by(pan_day) %>% 
  summarise(n = n(), 
            min = min(age),
            max = max(age), 
            mean = mean(age), 
            sd = sd(age))
head(age_stratified_after_panday_each, n = 4L)
```
 total: <em>57</em> observations of <em>6</em> variables


- for all days after start of pandemic combined
```{r pan_day_all, echo=FALSE}
age_stratified_after_panday_total <- data_tidier %>% 
  filter(pan_day > 50) %>% 
  summarise(n = n(), 
            min = min(age),
            max = max(age), 
            mean = mean(age), 
            sd = sd(age))
head(age_stratified_after_panday_total)
```
            
##### Fatemeh

##### Only for persons with `ct_result` == <em>45</em>
Calculating and outputting min, max, mean and sd of column `age` after grouping by `ct_result` == <em>45</em>.

```{r ct_result_all, echo=FALSE}
age_stratified_after_ctresult_total <- data_tidier %>%
  filter(ct_result == 45) %>%                               
  group_by(ct_result) %>% 
  summarise(n = n(), 
            min = min(age),
            max = max(age), 
            mean = mean(age), 
            sd = sd(age))
head(age_stratified_after_panday_total)
```


##### Pratik

##### Only for persons with patient_class == inpatient
Calculating and outputting min, max, mean and sd of column `pan_day` after filtering by `patient_class` == `inpatient`  and grouping by `gender`.

```{r patient_class_all, echo=FALSE}
age_stratified_after_patient_class_total <- data_tidier %>%
  filter(patient_class == "inpatient") %>% 
  group_by(gender) %>%
  summarise(
    min_value  = min(pan_day,na.rm = TRUE),
    max_value  = max(pan_day,na.rm = TRUE),
    mean_value = mean(pan_day,na.rm = TRUE),
    sd_value   = sd(pan_day,na.rm = TRUE),
    .groups = "drop"
    )
head(age_stratified_after_patient_class_total)
```

### Task 4 - Create plots that would help answer these questions:

#### Sebastian
#### Are there any correlated measurements?
```{r 4_correlation, echo=FALSE}
data <- data_tidier %>% 
  select(-c(ID))

x <- (ggcorr(
  data,
  method = c("complete.obs", "pearson"),
  name = "Legend",
  geom = "circle",
  max_size = 10,
  min_size = 1,
  nbreaks = 6,
  size = 2.5,
  hjust = 0.75,
  angle = 0,
  palette = "RdBu"
)) 
x
```

With a correlation matrix, we could observe some few correlations: </br>
 - Strongest correlations <em>(0,67 - 1)</em> between `pan_day` & `pan_week` and `orderset` & `multi_ct_result_orderset`. Here, correlation is expected since `pan_week` and `multi_ct_result_orderset` were created out of `pan_day` and `orderset`</br>
 - weaker correlations <em>(0,33 - 0,67)</em> between `age` & `pan_day`, `age` & `pan_week`, `age` & `col_rec_tat`. Again, `pan_week` is based on `pan_day` and is ignored</br>
 - rest mostly correlations between <em>0 - 0,33</em> and some between <em>-0,33 - 0</em> and not significant</br>

#### Pratik
#### Does the time spent waiting for the test result improve during these 100 days that the dataset includes?
```{r 4_plot1_P, echo=FALSE}
data_tidier%>%
  ggplot(aes(x = pan_day, y = rec_ver_tat)) +
  geom_point(alpha = 0.4, color = "steelblue") +     # scatterplot of points
  geom_smooth(method = "loess", se = TRUE, color = "red") +  # smooth trend line
  labs(
    title = "Change in waiting time for test results over 100 days",
    x = "Day after start of pandemic",
    y = "Waiting time (receive → verification)"
  ) +
  theme_minimal()
```

We tried to find a correlation between `rec_ver_tat` and `pan_day`. We cannot find improvements and the correlation matrix shows almost no correlation (r=0.01). Similarly, a scatterplot between these two variables also shows a flat which indicates no relationship between these two variables. 

#### Were there more females than males that took the test at a drive through?
```{r 4_table_P, echo=FALSE}
data_tidier %>%
  with(table(gender, drive_thru_ind))
```

Yes, there were more females <em>(n=3997)</em> than males <em>(n=3912)</em> to take test at a drive through. First, filter was done to select only cases with test done at a drive through and a table was generated to segregate by gender.

#### Fatemeh
#### Does the distribution of the `ct_result` differ with sex group?
```{r 4_plot1_F, echo=FALSE}
ggplot(data, aes(x = gender, y = ct_result, fill = gender)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.1, fill = "white") +
  labs(title = "Violin plot of ct_result by gender",
       x = "gender",
       y = "ct_result") +
  theme_minimal() +
  theme(legend.position = "none")
```

The violin plot indicates that the `ct_result` distributions for females and males are nearly identical, showing no meaningful difference between the two sex groups. Their medians and overall spread are very similar, with no distinct peaks. Therefore, `gender` has no meaningful effect on `ct_result`.

```{r 4_plot2_F, echo=FALSE}
ggplot(data, aes(x = ct_result, colour = gender, fill = gender)) +
  geom_density(alpha = 0.4) +
  labs(title = "Density of ct_result by gender",
       x = "ct_result",
       y = "Density") +
  theme_minimal()
```

Next, we tried a density plot. This plot shows that the `ct_result` distributions for females and males almost completely overlap, showing no meaningful difference between the two sex groups.

#### Sebastian
#### Does the distribution of the `ct_result` differ with payor_group?
```{r 4_plot1_S, echo=FALSE}
ggplot(data, aes(x = payor_group, y = ct_result, fill = payor_group)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.2, fill = "white") +
  labs(title = "Violin plot of CT results grouped by health care category",
       x = "health care",
       y = "ct result") +
  theme_minimal() +
  theme(legend.position = "none") +
  ylim(0, 50)
```

```{r 4_plot2_S, echo=FALSE}
ggplot(data, aes(y = payor_group, x = ct_result, fill = payor_group, height = after_stat(density))) +
  geom_density_ridges(scale = 2, stat = "density") +
  scale_y_discrete(expand = c(0.01, 0)) +
  theme_ridges() +
  labs(title = "Ridgeline chart of CT results grouped by health care category",
       x = "ct results",
       y = "") +
  theme(legend.position = "none") +
  xlim(10, 50)
```

For this question we used first a violin plot for the distribution shape and additionally a Ridgeline chart to show the density distribution for all groups. All `payor_group` share the largest distribution at `ct_result` = <em>45</em>, with `commercial` having the most results at <em>45</em>. `Government` and `medical assistance` have second highest number at <em>45</em>. Rest of the groups have approx. the same number. Some outlier are obserable in all groups, with `commercial`, `government` and `NA` having the most. Least outliers are in groups `charity care` and `other`, `medical assistnance`.


### Task 5 - Analyse the dataset and answer the following questions:

#### Pratik
#### Are there more positive tests in the drive through?
```{r 5_plot1_P, echo=FALSE}
data %>%
  with(table(result, drive_thru_ind))
data %>%
  with(prop.table(table(result, drive_thru_ind)))
```

When we see the outcome of the tests across the site of specimen collection, we find more positive tests in the drive-through group both by count <em>(479 versus 381)</em> and proportion <em>(3.1% versus 2.5%)</em>.

#### Fatemeh
#### Is there a difference in the distributions of `ct_results` between different outcome groups (result)?
```{r 5_plot1_F, echo=FALSE}
ggplot(data, aes(x = result, y = ct_result, fill = result)) +
  geom_boxplot(alpha = 0.7, outlier.color = "red") +
  labs(title = "Distribution of ct_result across outcome groups",
       x = "Outcome group (result)",
       y = "ct_result") +
  theme_minimal() +
  theme(legend.position = "none")
```

Yes, `ct_result` values are lower in `positive` cases and higher in `negative` ones, with no data for the `invalid` group.
The boxplot shows clear differences in `ct_result` by outcome group: `positives` have lower, more variable Ct values, while `negatives` and invalids cluster at higher, consistent levels, reflecting the expected inverse link between target amount and Ct.

#### Sebastian
#### Is the age of the individual associated with whether the test result is negative or positive?
```{r 5_plot1_S, echo=FALSE}
ggplot(data, aes(x = result, y = age, fill = result)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.2, fill = "white") +
  labs(title = "Violin plot representing the age distribution of a corona test result",
       x = "",
       y = "age") +
  theme_minimal() +
  theme(legend.position = "none")
```
  
Yes, in general we can say: the younger the person, the more likely it is for the result to be `negative`. Around <em>50 %</em> of individuals age <em>10</em> or younger have `negative` results. Second most distribution <em>(ca. 25 %)</em> age <em>10 - ~23</em>. The median is at age <em>10</em>.
In comparison, <em>50 %</em> of individuals age <em>20</em> or younger have `positive` results. Second most distribution <em>(ca. 25 %)</em> age <em>20 - ~27</em>. Median is at age <em>20</em>.
Individuals with `negative` results are approx. <em>10</em> years younger and mor common than individuals with `positive` result.


### Task 6 - Write a short report in an Rmd format.

We summarized our structure, data sets, procedures for every task and plots, including analysis in a RMarkdown protocol `exam_summary_data_freak.Rmd`. This file is located in the script/task6 folder and the .html output in the results folder.  


